---
title: "IPTW"
author: "Hai Nguyen"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = paste0(
        xfun::sans_ext(input), '-', Sys.Date(), '.html'
      ),
      output_dir = "../Output"
    )
  })
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    #code_folding: hide
    smart: true
    theme: united
    highlight: kate
    df_print: paged
---
## To do

  - To conduct an IPTW

```{r}
#install.packages("geepack") # install package if required
library("geepack")
```


```{r}
hr.b <- readRDS("../Data/HR_ready4analyze.Rds")
# names(hr.b)
# str(hr.b)
table(hr.b$race.ethnicity, useNA = "always")
sum(is.na(hr.b$age))
sum(is.na(hr.b$bmi))
table(hr.b$loc, useNA = "always")
```

## Modeling PS

  *... one can fit a very complicated propensity score model with interactions and higher order terms first. Since the goal of this propensity score model is to obtain the best estimated probability of treatment assignment, one is not concerned with over-parameterizing this model.*
  -- D'Agostino (quoting Rosenbaum and Rubin); D’agostino, R.B. 1998. Propensity score matching for bias reduction in the comparison of a treatment to a non-randomized control group. Statistical Medicine 17: 2265-2281.
  
```{r}
# Estimation of ip weights via a logistic model
fit <- glm(
  chd ~ as.factor(loc) + female + as.factor(race.ethnicity) + age + I(age^2) +
    bmi + I(bmi^2),
  family = binomial(),
  data = hr.b
)
summary(fit)


hr.b$race.ethnicity <- ifelse(is.na(hr.b$race.ethnicity), "unknown", hr.b$race.ethnicity)
table(hr.b$race.ethnicity, useNA = "always")
hr.b$race.ethnicity.miss <- ifelse(hr.b$race.ethnicity == "unknown", 1, 0)
table(hr.b$race.ethnicity.miss, useNA = "always")
fit <- glm(
  chd ~ as.factor(loc) + female + as.factor(race.ethnicity) + age + I(age^2) +
    bmi + I(bmi^2) + race.ethnicity.miss,
  family = binomial(),
  data = hr.b
)
summary(fit)
# hr.b$white.miss <- ifelse(hr.b$white == '.', 1, 0)
# hr.b$hispanic.miss <- ifelse(hr.b$hispanic == '.', 1, 0)
# hr.b$white <- ifelse(is.na(hr.b$white), 2, hr.b$white)
# 
# hr.b$hispanic <- ifelse(is.na(hr.b$hispanic), 2, hr.b$hispanic)
# 
# table(hr.b$white.miss, useNA = "always")
# table(hr.b$hispanic.miss, useNA = "always")
# table(hr.b$white, useNA = "always")
# table(hr.b$white, useNA = "always")
# 
# fit <- glm(
#   chd ~ as.factor(white) + as.factor(hispanic) + white.miss + hispanic.miss + 
#     female + 
#     age + I(age^2) +
#     bmi + I(bmi^2),
#   family = binomial(),
#   data = hr.b
# )
# summary(fit)

```

```{r}
z<-hr.b$chd
zf<-factor(z,levels=c(1,0),labels=c("chd","control"))
table(zf)
ps <- fit$fitted.values
hr.b$ps<-ps
boxplot(ps~zf, ylab="Propensity score", main="Exercise Cohort")
hr.b %>%
  ggplot(aes(x=ps, fill=factor(chd))) +
    geom_boxplot()
library(tidyverse)
library(hrbrthemes) #theme_ipsum
hr.b %>%
  ggplot(aes(x=ps, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    
    theme_ipsum() +
    labs(fill="")
```

```{r}
ggplot(hr.b, aes(x = ps, fill = as.factor(chd), color = as.factor(chd))) +
  geom_histogram(alpha = 0.3, position = 'identity', bins=15) +
  facet_grid(as.factor(chd) ~ .) +
  xlab('Probability of CHD') +
  ggtitle('Propensity Score Distribution by CHD Group') +
  scale_fill_discrete('') +
  scale_color_discrete('') +
  theme(legend.position = 'bottom', legend.direction = 'vertical')
```

```{r}
# attempt to reproduce plot from the book
# gg<-
hr.b %>%
  mutate(ps.grp = round(ps/0.05) * 0.05) %>%
  group_by(chd, ps.grp) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(n2 = ifelse(chd == 0, yes = n, no =  -1*n)) %>%
  ggplot(aes(x = ps.grp, y = n2, fill = as.factor(chd))) +
  geom_bar(stat = 'identity', position = 'identity') +
  geom_text(aes(label = n, x = ps.grp, y = n2 + ifelse(chd == 0, 8, -8))) +
  xlab('Probability of CHD') +
  ylab('N') +
  ggtitle('Propensity Score Distribution by CHD-Control Groups') +
  scale_fill_discrete('') +
  scale_x_continuous(breaks = seq(0, 1, 0.05)) +
  theme(legend.position = 'bottom', legend.direction = 'vertical',
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank())

# ggsave(filename = "PS Dist by Grps-CHD dataset.PNG", gg, width = 8, height = 5, dpi = 300, units = "in", device = "png")
```


## Estimating non-stabilized IP weights
```{r}
p.chd.obs <-
  ifelse(hr.b$chd == 0,
         1 - predict(fit, type = "response"),
         predict(fit, type = "response"))
```

```{r}
hr.b$p.chd.obs <- p.chd.obs
hr.b$wATE <- 1 / p.chd.obs
summary(hr.b$wATE)
sd(hr.b$wATE)
```

```{r}
hr.b$wATT <- hr.b$chd+p.chd.obs*(1-hr.b$chd)/(1-p.chd.obs)
summary(hr.b$wATT)
sd(hr.b$wATT)
```

```{r}
msm.lm <- lm(chronotropic.index ~ chd, data = hr.b)
summary(msm.lm)

msm.wATE <- geeglm(
  chronotropic.index ~ chd,
  data = hr.b,
  weights = wATE,
  id = id,
  corstr = "independence"
)
summary(msm.wATE)

msm.wATT <- geeglm(
  chronotropic.index ~ chd,
  data = hr.b,
  weights = wATT,
  id = id,
  corstr = "independence"
)
summary(msm.wATT)
beta <- coef(msm.wATT)
SE <- coef(summary(msm.wATT))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```

## Estimating stabilized IP weights

```{r}
# estimation of denominator of ip weights
denom.fit <- 
  glm(
    chd ~ as.factor(loc) + female + as.factor(race.ethnicity) + age + I(age^2) +
    bmi + I(bmi^2) + race.ethnicity.miss,
  family = binomial(),
  data = hr.b
)
summary(denom.fit)
```

```{r}
pd.chd <- predict(denom.fit, type = "response")

# estimation of numerator of ip weights
numer.fit <- glm(chd ~ 1, family = binomial(), data = hr.b)
summary(numer.fit)
```

```{r}
pn.chd <- predict(numer.fit, type = "response")

hr.b$sw <-
  ifelse(hr.b$chd == 0, ((1 - pn.chd) / (1 - pd.chd)),
         (pn.chd / pd.chd))

summary(hr.b$sw)
```
```{r}
msm.sw <- geeglm(
  chronotropic.index ~ chd,
  data = hr.b,
  weights = sw,
  id = id,
  corstr = "independence"
)
summary(msm.sw)
```

### Resting HR
```{r}
msm.sw <- geeglm(
  m0 ~ chd,
  data = hr.b,
  weights = sw,
  id = id,
  corstr = "independence"
)
summary(msm.sw)
beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```

### HR peak
```{r}
msm.sw <- geeglm(
  hr.max.achieved ~ chd,
  data = hr.b,
  weights = sw,
  id = id,
  corstr = "independence"
)
summary(msm.sw)

beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```


### HR reserve
```{r}
hr.b$hr.res <- hr.b$hr.max.achieved - hr.b$m0
msm.sw <- geeglm(
  hr.res ~ chd,
  data = hr.b,
  weights = sw,
  id = id,
  corstr = "independence"
)
summary(msm.sw)

beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```


### HR recovery

```{r}
hr.b$hr.rec1 <- hr.b$hr.max.achieved - hr.b$r1
hr.b$hr.rec2 <- hr.b$hr.max.achieved - hr.b$r2
hr.b$hr.rec3 <- hr.b$hr.max.achieved - hr.b$r3

# msm.sw <- geeglm(
#   hr.rec1 ~ chd,
#   data = hr.b,
#   weights = sw,
#   id = id,
#   corstr = "independence"
# )
# summary(msm.sw)
# 
# beta <- coef(msm.sw)
# SE <- coef(summary(msm.sw))[, 2]
# lcl <- beta - qnorm(0.975) * SE
# ucl <- beta + qnorm(0.975) * SE
# cbind(beta, lcl, ucl)


# msm.sw <- geeglm(
#   hr.rec2 ~ chd,
#   data = hr.b,
#   weights = sw,
#   id = id,
#   corstr = "independence"
# )
# summary(msm.sw)
# 
# beta <- coef(msm.sw)
# SE <- coef(summary(msm.sw))[, 2]
# lcl <- beta - qnorm(0.975) * SE
# ucl <- beta + qnorm(0.975) * SE
# cbind(beta, lcl, ucl)


msm.sw <- geeglm(
  hr.rec3 ~ chd,
  data = hr.b,
  weights = sw,
  id = id,
  corstr = "independence"
)
summary(msm.sw)

beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```

### Relative heart rate recovery
```{r}
hr.b$rhrr1 <- hr.b$r1/hr.b$hr.max.achieved
hr.b$rhrr2 <- hr.b$r2/hr.b$hr.max.achieved
hr.b$rhrr3 <- hr.b$r3/hr.b$hr.max.achieved
msm.sw <- geeglm(
  rhrr3 ~ chd,
  data = hr.b,
  weights = sw,
  id = id,
  corstr = "independence"
)
summary(msm.sw)

beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```

### Chronotropic index
```{r}
msm.sw <- geeglm(
  chronotropic.index ~ chd,
  data = hr.b,
  weights = sw,
  id = id,
  corstr = "independence"
)
summary(msm.sw)

beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```

### Chronotropic index
```{r}
msm.sw <- geeglm(
  vo2 ~ chd,
  data = hr.b,
  weights = sw,
  id = id,
  corstr = "independence"
)
summary(msm.sw)

beta <- coef(msm.sw)
SE <- coef(summary(msm.sw))[, 2]
lcl <- beta - qnorm(0.975) * SE
ucl <- beta + qnorm(0.975) * SE
cbind(beta, lcl, ucl)
```

