---
title: "Matching"
author: "Hai Nguyen"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = paste0(
        xfun::sans_ext(input), '-', Sys.Date(), '.html'
      ),
      output_dir = "../Output"
    )
  })
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    #code_folding: hide
    smart: true
    indent: true
    theme: united
    highlight: kate
    df_print: paged
---



## To-do

- Propensity Score Matching  on 4 factors: `loc`, `age`, `bmi` and `female`

## Load packages and data
```{r load packages, warning=F, message=F}
library(readxl) # read Excel
library(DT)## column filters in DT
library(psych) # describe by group
library(tidyverse)
library(hrbrthemes) #theme_ipsum
library(kableExtra)
library(reshape2) #melt
library(tableone) #CreateTableOne
library(DOS2)
library(optmatch)
hr.b <- readRDS(file = "../Data/HR_ready4analyze.Rds")

```

## Propensity score

The propensity score is estimated using a logit model. I decided to include those variables in PS model:  
  - `age`   
  - `bmi`  
  - `female`   
  - `white`  
  - `hispanic`   
  - **Not** put `vo2` in PS model   
  
```{r ps}
# Estimate the propensity score
z<-hr.b$chd
zf<-factor(z,levels=c(1,0),labels=c("chd","control"))
table(zf)

X<-hr.b[,c("white","hispanic","age","bmi","female")]

#hr.b$race.ethnicity <- ifelse(is.na(hr.b$race.ethnicity), "unknown", hr.b$race.ethnicity)
#hr.b$race.ethnicity.miss <- ifelse(hr.b$race.ethnicity == "unknown", 1, 0)
hr.b$white <- ifelse(is.na(hr.b$white), 2, hr.b$white)
hr.b$white.miss <- ifelse(hr.b$white == '.', 1, 0)
hr.b$hispanic <- ifelse(is.na(hr.b$hispanic), 2, hr.b$hispanic)
hr.b$hispanic.miss <- ifelse(hr.b$hispanic == '.', 1, 0)

fit <- glm(
  chd ~ as.factor(white) + as.factor(hispanic) + white.miss + hispanic.miss + 
    female + 
    age + I(age^2) +
    bmi + I(bmi^2),
  family = binomial(),
  data = hr.b
)
summary(fit)
ps <- fit$fitted.values
hr.b$ps<-ps
```

```{r boxplot}
boxplot(ps~zf, ylab="Propensity score", main="Exercise Cohort")
```
```{r}
hr.b$CHD <- as.factor(hr.b$chd)
ggplot(hr.b, aes(x = ps, fill = CHD)) + geom_density(alpha = 0.5)
```

## Table 1 before Matching
```{r}
xvars <- c("white","hispanic","age","bmi","female","ps")
hr.table1 <- hr.b
hr.table1$ps <- ps
#look at a table 1
table1<- CreateTableOne(vars=xvars,strata="chd", data=hr.table1, test=FALSE)
## include standardized mean difference (SMD)
print(table1,smd=TRUE)
```

## Matching 1:1
```{r smahal}
#hr.b$race.ethnicity <- as.numeric(as.factor(hr.b$race.ethnicity))
X<-hr.b[ , c("white","hispanic","age","bmi","female")]
#Robust Mahalanobis distance matrix, treated x control
dmat<-smahal(z,X)
dim(dmat)

kable(round(dmat[1:5,1:5],2), caption="First five rows and columns of the 145×320 distance matrix using the rank-based Mahalanobis distance") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

```{r}
#Add a caliper (.2xsd(PS))on the propensity score using a penalty function
dmat <- addcaliper(dmat, z, ps, caliper=.2)
```

```{r}
pm <- optmatch::pairmatch(dmat,data=hr.b)
```

```{r}
# Housekeepers
ipm <- as.integer(pm)
hr.pm <- cbind(hr.b, matches=pm,ipm)
df.pm<-hr.pm[matched(pm),] # only select matched cases

chd.data <- df.pm %>%
  filter(chd==1) %>%
  arrange(ipm)

matched.controls <- df.pm %>%
  filter(chd==0) %>%
  arrange(ipm)

table1.matched<- CreateTableOne(vars=xvars,strata="chd", data=df.pm, test=FALSE)
## include standardized mean difference (SMD)
print(table1.matched, smd=TRUE)
```

```{r}
# alternative plot with histograms
df.pm <- df.pm %>% mutate(chdlabel = ifelse(chd == 1,
                       yes = 'CHD',
                       no = 'Controls'))
ggplot(df.pm, aes(x = ps, fill = as.factor(chd), color = as.factor(chd))) +
  geom_histogram(alpha = 0.3, position = 'identity', bins=15) +
  facet_grid(as.factor(chd) ~ .) +
  xlab('Probability of CHD') +
  ggtitle('Propensity Score Distribution by CHD Group') +
  scale_fill_discrete('') +
  scale_color_discrete('') +
  theme(legend.position = 'bottom', legend.direction = 'vertical')

# attempt to reproduce plot from the book
#gg<-
df.pm %>%
  mutate(ps.grp = round(ps/0.05) * 0.05) %>%
  group_by(chd, ps.grp) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(n2 = ifelse(chd == 0, yes = n, no =  -1*n)) %>%
  ggplot(aes(x = ps.grp, y = n2, fill = as.factor(chd))) +
  geom_bar(stat = 'identity', position = 'identity') +
  geom_text(aes(label = n, x = ps.grp, y = n2 + ifelse(chd == 0, 8, -8))) +
  xlab('Probability of CHD') +
  ylab('N') +
  ggtitle('Propensity Score Distribution by CHD-Control Groups on Optimal Pair Match') +
  scale_fill_discrete('') +
  scale_x_continuous(breaks = seq(0, 1, 0.05)) +
  theme(legend.position = 'bottom', legend.direction = 'vertical',
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank())

#ggsave(filename = "PS Dist by Grps on Optimal Pair Match.PNG", gg, width = 8, height = 5, dpi = 300, units = "in", device = "png")
```


### Write matched data out
```{r, eval=FALSE}
## write to Excel (only execute one)
openxlsx::write.xlsx(df.pm, "../data/Matched_1-1_HRExercise.xlsx")
saveRDS(df.pm, file="../data/Matched_1-1_HRExercise.Rds")
```

## Compare

### Resting HR
```{r hr rest, warning=F, message=F}
describe(df.pm$m0)
describeBy(df.pm$m0, group=df.pm$chd)
df.pm %>%
  ggplot(aes(x=m0, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")
boxplot(df.pm$m0 ~ df.pm$chd,ylab="HR Rest",xlab="CHD",main="Exercise Study")
t.test(df.pm$m0 ~ df.pm$chd)
wilcox.test(df.pm$m0 ~ df.pm$chd)
```

### HR peak
```{r hr peak, warning=F, message=F}
describe(df.pm$hr.max.achieved)
describeBy(df.pm$hr.max.achieved, group=df.pm$chd)
df.pm %>%
  ggplot(aes(x=hr.max.achieved, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")
boxplot(df.pm$hr.max.achieved ~ df.pm$chd,ylab="HR Peak", xlab="CHD", main="Exercise Study")
t.test(df.pm$hr.max.achieved ~ df.pm$chd)
wilcox.test(df.pm$hr.max.achieved ~ df.pm$chd)
```

### HR reserve
```{r hr res, warning=F, message=F}
df.pm$hr.res <- df.pm$hr.max.achieved - df.pm$m0

describe(df.pm$hr.res)
describeBy(df.pm$hr.res, group=df.pm$chd)

df.pm %>%
  ggplot(aes(x=hr.res, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")
boxplot(df.pm$hr.res ~ df.pm$chd,ylab="HR Reserve",xlab="CHD", main="Exercise Study")
t.test(df.pm$hr.res ~ df.pm$chd)
wilcox.test(df.pm$hr.res ~ df.pm$chd)
```

### HR recovery
```{r hr rec, warning=F, message=F}
df.pm$hr.rec1 <- df.pm$hr.max.achieved - df.pm$r1
df.pm$hr.rec2 <- df.pm$hr.max.achieved - df.pm$r2
df.pm$hr.rec3 <- df.pm$hr.max.achieved - df.pm$r3
df.pm$hr.rec4 <- df.pm$hr.max.achieved - df.pm$r4
df.pm$hr.rec5 <- df.pm$hr.max.achieved - df.pm$r5
df.pm$hr.rec6 <- df.pm$hr.max.achieved - df.pm$r6

describe(df.pm$hr.rec1)
describeBy(df.pm$hr.rec1, group=df.pm$chd)
df.pm %>%
  ggplot(aes(x=hr.rec1, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")
boxplot(df.pm$hr.rec1 ~ df.pm$chd,ylab="HR Recover 1m",xlab="CHD", main="Exercise Study")
t.test(df.pm$hr.rec1 ~ df.pm$chd)
wilcox.test(df.pm$hr.rec1 ~ df.pm$chd)

describe(df.pm$hr.rec2)
describeBy(df.pm$hr.rec2, group=df.pm$chd)
df.pm %>%
  ggplot(aes(x=hr.rec2, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")
boxplot(df.pm$hr.rec2 ~ df.pm$chd,ylab="HR Recover 2m",xlab="CHD", main="Exercise Study")
t.test(df.pm$hr.rec2 ~ df.pm$chd)
wilcox.test(df.pm$hr.rec2 ~ df.pm$chd)

describe(df.pm$hr.rec3)
describeBy(df.pm$hr.rec3, group=df.pm$chd)
df.pm %>%
  ggplot(aes(x=hr.rec3, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")
boxplot(df.pm$hr.rec3 ~ df.pm$chd,ylab="HR Recover 3m",xlab="CHD", main="Exercise Study")
t.test(df.pm$hr.rec3 ~ df.pm$chd)
wilcox.test(df.pm$hr.rec3 ~ df.pm$chd)

describe(df.pm$hr.rec4)
describeBy(df.pm$hr.rec4, group=df.pm$chd)
df.pm %>%
  ggplot(aes(x=hr.rec4, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")
boxplot(df.pm$hr.rec4 ~ df.pm$chd, ylab="HR Recover 4m", xlab="CHD", main="Exercise Study")
t.test(df.pm$hr.rec4 ~ df.pm$chd)
wilcox.test(df.pm$hr.rec4 ~ df.pm$chd)

describe(df.pm$hr.rec5)
describeBy(df.pm$hr.rec5, group=df.pm$chd)
df.pm %>%
  ggplot(aes(x=hr.rec5, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")
boxplot(df.pm$hr.rec5 ~ df.pm$chd,ylab="HR Recover 5m",xlab="CHD", main="Exercise Study")
t.test(df.pm$hr.rec5 ~ df.pm$chd)
wilcox.test(df.pm$hr.rec5 ~ df.pm$chd)

describe(df.pm$hr.rec6)
describeBy(df.pm$hr.rec6, group=df.pm$chd)
df.pm %>%
  ggplot(aes(x=hr.rec6, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")
boxplot(df.pm$hr.rec6 ~ df.pm$chd,ylab="HR Recover 6m",xlab="CHD", main="Exercise Study")
t.test(df.pm$hr.rec6 ~ df.pm$chd)
wilcox.test(df.pm$hr.rec6 ~ df.pm$chd)
```


### Relative heart rate recovery
```{r rhrr manipulate, warning=F, message=F}
# Relative heart rate recovery: ratio of heart rate in recovery to the peak heart rate at each minute 
df.pm$rhrr1 <- df.pm$r1/df.pm$hr.max.achieved
df.pm$rhrr2 <- df.pm$r2/df.pm$hr.max.achieved
df.pm$rhrr3 <- df.pm$r3/df.pm$hr.max.achieved
df.pm$rhrr4 <- df.pm$r4/df.pm$hr.max.achieved
df.pm$rhrr5 <- df.pm$r5/df.pm$hr.max.achieved
df.pm$rhrr6 <- df.pm$r6/df.pm$hr.max.achieved


tab.rhrr <- df.pm %>%
  group_by(chd) %>%
  summarise_at(vars("rhrr1":"rhrr6"), funs(mean(., na.rm=TRUE), sd(., na.rm=TRUE)))
round(tab.rhrr*100, 1)

df.pm.rhrr <- df.pm %>%
  select(id, serNum, chd, rhrr1, rhrr2, rhrr3, rhrr4, rhrr5, rhrr6)

df.pm.rhrr.long <- melt(df.pm.rhrr, id.vars = c("id","serNum", "chd"),
    variable.name = "min", value.name = "rhrr")


boxplot(rhrr ~ chd, ylab="Relative HR Recovery 1m",xlab="CHD",main="Exercise Study", data = df.pm.rhrr.long, subset = min=="rhrr1")
t.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr1")
wilcox.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr1")

boxplot(rhrr ~ chd, ylab="Relative HR Recovery 1m",xlab="CHD",main="Exercise Study", data = df.pm.rhrr.long, subset = min=="rhrr2")
t.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr2")
wilcox.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr2")

boxplot(rhrr ~ chd, ylab="Relative HR Recovery 1m",xlab="CHD",main="Exercise Study", data = df.pm.rhrr.long, subset = min=="rhrr3")
t.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr3")
wilcox.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr3")

boxplot(rhrr ~ chd, ylab="Relative HR Recovery 1m",xlab="CHD",main="Exercise Study", data = df.pm.rhrr.long, subset = min=="rhrr4")
t.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr4")
wilcox.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr4")

boxplot(rhrr ~ chd, ylab="Relative HR Recovery 1m",xlab="CHD",main="Exercise Study", data = df.pm.rhrr.long, subset = min=="rhrr5")
t.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr5")
wilcox.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr5")

boxplot(rhrr ~ chd, ylab="Relative HR Recovery 1m",xlab="CHD",main="Exercise Study", data = df.pm.rhrr.long, subset = min=="rhrr6")
t.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr6")
wilcox.test(rhrr ~ chd, data = df.pm.rhrr.long, subset = min=="rhrr6")
```


### Chronotropic index

```{r Chronotropic index, warning=F, message=F}
describe(df.pm$chronotropic.index)
describeBy(df.pm$chronotropic.index, group=df.pm$chd)

df.pm %>%
  ggplot(aes(x=chronotropic.index, fill=factor(chd))) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_ipsum() +
    labs(fill="")

boxplot(df.pm$chronotropic.index ~ df.pm$chd,ylab="Chronotropic Index", xlab="CHD", main="Exercise Study")
t.test(df.pm$chronotropic.index ~ df.pm$chd)
wilcox.test(df.pm$chronotropic.index ~ df.pm$chd)
```

```{r}
t.test(df.pm$vo2 ~ df.pm$chd)
```

